Sparse spectral methods for partial differential equations on spherical caps

### 2020-12-21 papers/spherical-caps/spherical-caps.tex
Fixed image sizing
final(?)

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, disk-slices and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and Biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere -- specifically spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
where $\alpha \in (-1,1)$ and $\beta := 1$.
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.

@@

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere. This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients. Further, we remove the singular nature as $\alpha$ approaches $0$ that such a projection would give, since our new approach is applicable for all $\alpha \in (-1,1)$. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$ when $a > 0$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
\end{align*}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical \cite[\S5]{magnus1995painleve}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations.

@@

\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:construction}
	Let $w : (\alpha,\beta) \: \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: w(\cosphi) \: \sinphi \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \cos\theta, \rho(z) \sin\theta, z\big) \: q\big(\rho(z) \cos\theta, \rho(z) \sin \theta, z\big) \: w(z) \: \D \theta \: \D z
\end{align*}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

@@

The recurrences in Lemma \ref{lemma:Qrecurrence} lead to Jacobi operators that correspond to multiplication by $x$, $y$ and $z$. In later sections we will use an ordering of the OPs so that they are grouped by Fourier mode $k$, which is convenient for the application of differential and other operators to the vector of coefficients of a given function's expansion (some operators will exploit this ordering for operators where Fourier modes do not interact, and thus will be block-diagonal). Before that though, the ordering we will use in the remainder of this section is convenient for establishing Jacobi operators for multiplication by $x$, $y$ and $z$, and hence building the OPs and importantly obtaining the associated \textit{recurrence coefficient matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings. To this end, we define our OP-building ordering as follows. For $n=0,1,2,\dots$:
\begin{align*}
	\bigscoptna := 
		\begin{pmatrix}
			\scopa_{n,0,0}(x,y,z) \\
			\scopa_{n,1,0}(x,y,z) \\
			\scopa_{n,1,1}(x,y,z) \\
			\vdots \\
			\scopa_{n,n,0}(x,y,z) \\
			\scopa_{n,n,1}(x,y,z)
		\end{pmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigscopta := 
		\begin{pmatrix}
			\bigscopta_0 \\
			\bigscopta_1 \\
			\bigscopta_2 \\
			\vdots \\
		\end{pmatrix}
\end{align*}
and set $\tilde J_x^{(a)}, \tilde J_y^{(a)}, \tilde J_z^{(a)}$ as the Jacobi matrices corresponding to
\begin{align}
	\tilde J_x^{(a)} \: \bigscopta(x,y,z) = x \: \bigscopta(x,y,z), \nonumber \\
	\tilde J_y^{(a)} \: \bigscopta(x,y,z) = y \: \bigscopta(x,y,z), \label{eqn:jacobimatricesdefinitionalt} \\
	\tilde J_z^{(a)} \: \bigscopta(x,y,z) = z \: \bigscopta(x,y,z). \nonumber
\end{align}
where
\begin{align*}
	\tilde J_{x/y/z}^{(a)} &= 
		\begin{pmatrix}
			\tilde B^{(a)}_{x/y/z, 0} & \tilde A^{(a)}_{x/y/z, 0} & & & & \\
			\tilde C^{(a)}_{x/y/z, 1} & \tilde B^{(a)}_{x/y/z, 1} & \tilde A^{(a)}_{x/y/z, 1} & & & \\
			& \tilde C^{(a)}_{x/y/z, 2} & \tilde B^{(a)}_{x/y/z, 2} & \tilde A^{(a)}_{x/y/z, 2} & & & \\
			& & \tilde C^{(a)}_{x/y/z, 3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
\end{align*}
Note that $J_x^{(a)}, J_y^{(a)}, J_z^{(a)}$ are \textit{banded-block-banded matrices}:

@@

Each of these Jacobi matrices are then block-tridiagonal (block-bandwidths $(1,1)$). For $\tilde J_x^{(a)}$, the sub-blocks have sub-block-bandwidths $(2,2)$:

@@

	\tilde A^{(a)}_{x,n} &:= 

@@

	\tilde B^{(a)}_{x,n} &:= 

@@

			& \ddots & \ddots & \tilde A^{(a)}_{n,n-1,4} \\

@@

	\tilde C^{(a)}_{x,n} &:= 

@@

For $\tilde J_y^{(a)}$, the sub-blocks have sub-block-bandwidths $(3,3)$:

@@

	\tilde A^{(a)}_{y,n} &:= 

@@

	\tilde B^{(a)}_{y,n} &:= 

@@

	\tilde C^{(a)}_{y,n} &:= 

@@

For $\tilde J_z^{(a)}$, the sub-blocks are diagonal, i.e. have sub-block-bandwidths $(0,0)$:

@@

	\tilde A^{(a)}_{z,n} &:= 

@@

	\tilde B^{(a)}_{z,n} &:= 

@@

	\tilde C^{(a)}_{z,n} &:= 

@@

Following the triangle case \cite{olver2019triangle}, we can combine each system in (\ref{eqn:jacobimatricesdefinitionalt}) into a block-tridiagonal system, for any $(x,y,z) \in \Omega$:
\begin{align*}
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\bigscopta(x,y,z) =
		\begin{pmatrix}
	 		\scopa_0 \\ 0 \\ 0 \\ 0 \\ \vdots  \\
		\end{pmatrix},
\end{align*}
where we note $\scopa_0 := \scopa_{0,0,0}(x,y,z) \equiv \genjac_0^{(a,0)} \: \ch_0$, and for each $n = 0,1,2\dots$,
\begin{align*}
	A_n &:= 
		\begin{pmatrix}
			A^{(a)}_{x,n} \\
			A^{(a)}_{y,n} \\
			A^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+3)}, \quad
	C_n := 
		\begin{pmatrix}
			C^{(a)}_{x,n} \\
			C^{(a)}_{y,n} \\
			C^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n-1)} \quad (n \ne 0), \nonumber \\
	B_n &:= 
		\begin{pmatrix}
			B^{(a)}_{x,n} \\
			B^{(a)}_{y,n} \\
			B^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+1)}, \quad
	G_n(x,y,z) := 
		\begin{pmatrix}
			xI_{2n+1} \\
			yI_{2n+1} \\
			zI_{2n+1}
		\end{pmatrix} \in \R^{3(2n+1)\times(n+1)}.
\end{align*}

@@

\begin{theorem}\label{theorem:sparsityofdifferentialoperators}
	The operator matrices $D_\varphi^{(a)}, \: W_\varphi^{(a)}, \: D_\theta, \: \calL^{(a)\to(a+\tilde a)}, \: \calL_W^{(a)\to(a-\tilde a)}, \: \Delta^{(1)}_W$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure. More specifically:
\begin{itemize}
	\item $D_\varphi^{(a)}$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(2, 4)$
  	\item $W_\varphi^{(a)}$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(4, 2)$
	\item $D_\theta$ has  block-bandwidths $(0,0)$, and sub-block-bandwidths $(1, 1)$
	\item $\calL^{(a)\to(a+\tilde a)}$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(0, 4)$
	\item $\calL_W^{(a)\to(a-\tilde a)}$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(4, 0)$
	\item $\Delta^{(1)}_W$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(2, 2)$
\end{itemize}
\end{theorem}

@@

	&\quad \quad \quad \quad \cdot \: \Big( \int_\alpha^1 \: \genjacmmj^{(a - \tilde a, 2j)} \: \genjacw^{(a - 2, k+j)} \: \Big\{\genjacnmk^{(a,2k)} [-k(k+1) \genjacw^{(2,0)} - 2a(k+1)z \: \genjacw^{(1,0)} + a (a-1) \rho^2] \\

@@

	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + \genjacnmk^{(a,2k)} \: \genjacw^{(a-1, 2k)} [\genjac_{m-k}^{(a - \tilde a, 2k) \: \prime \prime} \: \genjacw^{(1, 2)} + a \genjac_{m-k}^{(a - \tilde a, 2k) \: \prime} \: \rho^2 - 2(k+1) z \: \genjac_{m-k}^{(a - \tilde a, 2k)} \: \genjacw^{(1, 0)}] \Big\} \: \D z \\

@@

	\medskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.23]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $\calB_W^{(2)}$} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$.}
	\label{fig:sparsity}
\end{figure}

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,u,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in Figure \ref{fig:sparsity}.

@@

The Clenshaw algorithm outlined in Section \ref{subsection:functionevaluation} can also be used with Jacobi matrices $J^{(a)}_x, J^{(a)}_y, J^{(a)}_z$ replacing the point $(x,y,z)$. Let $v : \Omega \to \R$ be the function that we wish to obtain an operator matrix $V$ for $v$, so that
\begin{align*}
	v(x,y,z) \: f(x,y,z) = v(x,y,z) \: \bm{f}^\top \: \bigscopNa(x,y,z) = (V \bm{f})^\top \: \bigscopNa(x,y,z),
\end{align*}
i.e. $V \bm{f}$ is the coefficients vector for the function $v(x,y,z) \: f(x,y,z)$.

>>>>

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, disk-slices and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere---specifically spherical caps. More precisely, we consider the solution of partial differential equations on the spherical cap $\Omega$ defined by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
where $\alpha \in (-1,1)$ and $\beta := 1$.

@@

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere. This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients. Further, we avoid the singular nature at the poles or  as $\alpha$ approaches $0$ that  such a projection may give, since our new approach yields a smooth polynomial basis for all $\alpha \in [-1,1)$.

On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \Omega$ when $a > 0$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
\end{align*}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical \cite[\S5]{magnus1995painleve}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals have already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations.

@@

\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:construction}
	Let $w : (\alpha,\beta) \: \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: \times \\
	&\qquad w(\cosphi) \: \sinphi \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \cos\theta, \rho(z) \sin\theta, z\big) \: q\big(\rho(z) \cos\theta, \rho(z) \sin \theta, z\big) \: w(z) \: \D \theta \: \D z
\end{align*}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

@@

The recurrences in Lemma \ref{lemma:Qrecurrence} lead to Jacobi operators that correspond to multiplication by $x$, $y$ and $z$. In later sections we will use an ordering of the OPs so that they are grouped by Fourier mode $k$, which is convenient for the application of differential and other operators to the vector of coefficients of a given function's expansion (some operators will exploit this ordering for operators where Fourier modes do not interact, and thus will be block-diagonal). Before that though, the ordering we will use in the remainder of this section is convenient for establishing Jacobi operators for multiplication by $x$, $y$ and $z$, and hence building the OPs and importantly obtaining the associated \textit{recurrence coefficient matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings. To this end, we define our OP-building ordering as follows. For $n=0,1,2,\dots$:
\begin{align*}
	\bigscoptna := 
		\begin{pmatrix}
			\scopa_{n,0,0}(x,y,z) \\
			\scopa_{n,1,0}(x,y,z) \\
			\scopa_{n,1,1}(x,y,z) \\
			\vdots \\
			\scopa_{n,n,0}(x,y,z) \\
			\scopa_{n,n,1}(x,y,z)
		\end{pmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigscopta := 
		\begin{pmatrix}
			\bigscopta_0 \\
			\bigscopta_1 \\
			\bigscopta_2 \\
			\vdots \\
		\end{pmatrix}
\end{align*}
and set $ J_x^{(a)},  J_y^{(a)},  J_z^{(a)}$ as the Jacobi matrices corresponding to
\begin{align}
	J_x^{(a)} \: \bigscopta(x,y,z) = x \: \bigscopta(x,y,z), \nonumber \\
	J_y^{(a)} \: \bigscopta(x,y,z) = y \: \bigscopta(x,y,z), \label{eqn:jacobimatricesdefinitionalt} \\
	J_z^{(a)} \: \bigscopta(x,y,z) = z \: \bigscopta(x,y,z). \nonumber
\end{align}
where
\begin{align*}
	J_{x/y/z}^{(a)} &= 
		\begin{pmatrix}
			B^{(a)}_{x/y/z, 0} & A^{(a)}_{x/y/z, 0} & & & & \\
			C^{(a)}_{x/y/z, 1} & B^{(a)}_{x/y/z, 1} & A^{(a)}_{x/y/z, 1} & & & \\
			& C^{(a)}_{x/y/z, 2} & B^{(a)}_{x/y/z, 2} & A^{(a)}_{x/y/z, 2} & & & \\
			& & C^{(a)}_{x/y/z, 3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
\end{align*}
Note that $J_x^{(a)}, J_y^{(a)}, J_z^{(a)}$ are \textit{banded-block-banded matrices}:

@@

Each of these Jacobi matrices are then block-tridiagonal (block-bandwidths $(1,1)$). For $J_x^{(a)}$, the sub-blocks have sub-block-bandwidths $(2,2)$:

@@

	A^{(a)}_{x,n} &:= 

@@

	B^{(a)}_{x,n} &:= 

@@

			& \ddots & \ddots & A^{(a)}_{n,n-1,4} \\

@@

	C^{(a)}_{x,n} &:= 

@@

For $J_y^{(a)}$, the sub-blocks have sub-block-bandwidths $(3,3)$:

@@

	A^{(a)}_{y,n} &:= 

@@

	B^{(a)}_{y,n} &:= 

@@

	C^{(a)}_{y,n} &:= 

@@

For $J_z^{(a)}$, the sub-blocks are diagonal, i.e. have sub-block-bandwidths $(0,0)$:

@@

	A^{(a)}_{z,n} &:= 

@@

	B^{(a)}_{z,n} &:= 

@@

	C^{(a)}_{z,n} &:= 

@@

Following the triangle case \cite{olver2019triangle}, we use a multivariate analogue of Clenshaw's algorithm for evaluation, where we  combine each system in (\ref{eqn:jacobimatricesdefinitionalt}) into a block-tridiagonal system, for any $(x,y,z) \in \Omega$:
\begin{align*}
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\bigscopta(x,y,z) =
		\begin{pmatrix}
	 		\scopa_0 \\ 0 \\ 0 \\ 0 \\ \vdots  \\
		\end{pmatrix},
\end{align*}
where we note $\scopa_0 := \scopa_{0,0,0}(x,y,z) \equiv \genjac_0^{(a,0)} \: \ch_0$, and for each $n = 0,1,2\dots$,
\begin{align*}
	A_n &:= 
		\begin{pmatrix}
			A^{(a)}_{x,n} \\
			A^{(a)}_{y,n} \\
			A^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+3)}, \quad
	C_n := 
		\begin{pmatrix}
			C^{(a)}_{x,n} \\
			C^{(a)}_{y,n} \\
			C^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n-1)} \quad (n \ne 0), \nonumber \\
	B_n &:= 
		\begin{pmatrix}
			B^{(a)}_{x,n} \\
			B^{(a)}_{y,n} \\
			B^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+1)}, \quad
	G_n(x,y,z) := 
		\begin{pmatrix}
			xI_{2n+1} \\
			yI_{2n+1} \\
			zI_{2n+1}
		\end{pmatrix} \in \R^{3(2n+1)\times(n+1)}.
\end{align*}

@@

\begin{theorem}\label{theorem:sparsityofdifferentialoperators}
	The operator matrices $D_\varphi^{(a)}, \: W_\varphi^{(a)}, \: D_\theta, \: \calL^{(a)\to(a+\tilde a)}, \: \calL_W^{(a)\to(a-\tilde a)}, \: \Delta^{(1)}_W$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure. More specifically:
\begin{itemize}
	\item $D_\varphi^{(a)}$ is block-diagonal with sub-block-bandwidths $(2, 4)$
  	\item $W_\varphi^{(a)}$ is block-diagonal with sub-block-bandwidths $(4, 2)$
	\item $D_\theta$ is block-diagonal with sub-block-bandwidths $(1, 1)$
	\item $\calL^{(a)\to(a+\tilde a)}$ is block-diagonal with sub-block-bandwidths $(0, 4)$
	\item $\calL_W^{(a)\to(a-\tilde a)}$ is block-diagonal with sub-block-bandwidths $(4, 0)$
	\item $\Delta^{(1)}_W$ is block-diagonal with sub-block-bandwidths $(2, 2)$
\end{itemize}
\end{theorem}

@@

	&\quad \quad  \cdot \: \Big( \int_\alpha^1 \: \genjacmmj^{(a - \tilde a, 2j)} \: \genjacw^{(a - 2, k+j)} \: \Big\{\genjacnmk^{(a,2k)} [-k(k+1) \genjacw^{(2,0)} - 2a(k+1)z \: \genjacw^{(1,0)} + a (a-1) \rho^2] \\

@@

	&\quad \quad \quad \quad \quad \quad \quad + a \: \genjacnmk^{(a,2k) \: \prime} \: \genjac_{m-k}^{(a - \tilde a, 2k)} \: \genjacw^{(a - 1, 2k+2)} \\
	&\quad \quad \quad \quad \quad \quad \quad + \genjacnmk^{(a,2k)} \: \genjacw^{(a-1, 2k)} [\genjac_{m-k}^{(a - \tilde a, 2k) \: \prime \prime} \: \genjacw^{(1, 2)} + a \genjac_{m-k}^{(a - \tilde a, 2k) \: \prime} \: \rho^2 - 2(k+1) z \: \genjac_{m-k}^{(a - \tilde a, 2k)} \: \genjacw^{(1, 0)}] \Big\} \: \D z \\

@@

	\medskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $\calB_W^{(2)}$} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$.}
	\label{fig:sparsity}
\end{figure}

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in Figure \ref{fig:sparsity}.

@@

The Clenshaw algorithm outlined in Section \ref{subsection:functionevaluation} can also be used with Jacobi matrices $J^{(a)}_x, J^{(a)}_y, J^{(a)}_z$ replacing the point $(x,y,z)$. Let $v : \Omega \to \R$ be the function that we wish to obtain an operator matrix $V$ for $v$, so that
\begin{align*}
	v(x,y,z) \: f(x,y,z) = v(x,y,z) \: \bm{f} \: \bigscopta(x,y,z) = (V \bm{f})^\top \: \bigscopta(x,y,z),
\end{align*}
i.e. $V \bm{f}$ is the coefficients vector for the function $v(x,y,z) \: f(x,y,z)$.
